\chapter*{Julian Tschannen}
\pagestyle{empty}

\vspace{-1cm}

\begin{tabbing}
Nationality: \hspace{1mm} \= Swiss \\
Date of birth: \> October 27, 1982 \\
Homepage:      \> \href{http://jt.x73.ch}{\texttt{jt.x73.ch}} \\
Email:         \> \href{mailto:jt@x73.ch}{\texttt{jt@x73.ch}} \\
\end{tabbing}


\section*{Work \& Education}

\begin{itemize}
  \item
  10/2009 -- 04/2015: \textbf{Dr. sc. ETH Zurich}, Switzerland\\
  Research assistant at the Chair of Software Engineering in the Department of Computer Science at ETH Zurich.

  \item
  10/2007 -- 04/2009: \textbf{MSc ETH in Computer Science}, Switzerland\\
  Degree with distinction, specialization in Software Engineering

  \item
  10/2006 -- 08/2007: \textbf{Internship at Eiffel Software}, USA

  \item
  10/2003 -- 10/2006: \textbf{BSc ETH in Computer Science}, Switzerland

\end{itemize}


\section*{Administrative \& Teaching Activities}

\begin{itemize}

  \item
  09/2010 -- 09/2014: \textbf{Organizer of LASER Summer School}\\
  I have co-organized the LASER summer school five times. The summer school is held in Elba, Italy.

  \item
  09/2012 -- 03/2015: \textbf{Project supervisor}\\
  I have supervised three lab projects, two Bachelor's theses, and one Master's thesis.

  \item
  09/2009 -- 12/2014: \textbf{Teaching assistant and Guest Lecturer}\\
  I have been head- or teaching assistant in nine courses and given various guest lectures.

\end{itemize}


\section*{Publications}

\begin{itemize}

  \item
  C. A. Furia, C. Poskitt, and J. Tschannen.
  \emph{The AutoProof Verifier: Usability by Non-Experts and on Standard Code}.
  In F-IDE, 2015.

  \item
  N. Polikarpova, J. Tschannen, and C. A. Furia.
  \emph{A Fully Verified Container Library}.
  In FM, 2015.

  \item
  J. Tschannen, C. A. Furia, M. Nordio, and N. Polikarpova. 
  \emph{AutoProof: Auto-active Functional Verification of Object-oriented Programs}.
  In TACAS, 2015

  \item
  N. Polikarpova, J. Tschannen, C. A. Furia, and B. Meyer.
  \emph{Flexible Invariants Through Semantic Collaboration}.
  In FM, 2014.

  \item
  J. Tschannen, C. A. Furia, and M. Nordio.
  \emph{AutoProof Meets Some Verification Challenges}.
  In STTT, 2014.

  \item
  J. Tschannen, C. A. Furia, M. Nordio, and B. Meyer.
  \emph{Program Checking With Less Hassle}.
  In VSTTE, 2013.

  \item
  J. Tschannen, C. A. Furia, M. Nordio, and B. Meyer.
  \emph{Automatic Verification of Advanced Object-Oriented Features: The AutoProof Approach}.
  In LASER: Tools for Practical Software Verification, 2012.

  \item
  T. Bormer, et al.
  \emph{The COST IC0701 Verification Competition 2011}.
  In FoVeOos, 2011.

  \item
  J. Tschannen, C. A. Furia, M. Nordio, and B. Meyer.
  \emph{Usable Verification of Object-Oriented Programs by Combining Static and Dynamic Techniques}.
  In SEFM, 2011.

  \item
  J. Tschannen, C. A. Furia, M. Nordio, and B. Meyer.
  \emph{Verifying Eiffel Programs with Boogie}.
  In BOOGIE, 2011.

  \item
  M. Nordio, H.-C. Estler, B. Meyer, J. Tschannen, C. Ghezzi, and E. Di Nitto.
  \emph{How do Distribution and Time Zones affect Software Development? A Case Study on Communication}.
  In ICGSE, 2011.

  \item
  M. Nordio, C. Ghezzi, B. Meyer, E. Di Nitto, G. Tamburrelli, J. Tschannen, N. Aguirre, and V. Kulkarni,
  \emph{Teaching Software Engineering using Globally Distributed Projects: the DOSE course}.
  In CTGDSD, 2011.

  \item
  M. Nordio, C. Calcagno, B. Meyer, P. M\"uller, and J. Tschannen.
  \emph{Reasoning about Function Objects}.
  In TOOLS-EUROPE, 2010.

\end{itemize}
