%============================================================================
\section{The Tools of \EVE} \label{tools-of-eve}
%============================================================================

The integrated verification techniques currently available in \EVE 
and partially illustrated in the preceding example session relies on several 
tools that are integrated in \EVE: \AutoProof, \AutoTest, and the \Inspector, which are now presented.

%----------------------------------------------------------------------------
\subsubsection{\AutoTest}
%----------------------------------------------------------------------------

\AutoTest~\cite{MEYER09}---now a standard component of commercial EiffelStudio---is a fully automatic contract-based testing tool.
\AutoTest generates objects by random calls to creation procedures.
Preconditions select valid inputs and postconditions serve as oracles: every test case consists of the execution of a routine on objects satisfying its precondition; if executing the routine violate its postcondition or calls another routine without satisfying its precondition, the routine tested has a fault.
\AutoTest is fully automatic and does not need any user input. Since contracts are used as test oracles, the tool can find functional correctness bugs.
A failing test case provides a concrete error report which is useful for debugging.

Like any dynamic technique based on execution, \AutoTest handles every feature of the source language (Eiffel).
Among its limitations, instead, is that random testing can take several hours to find the most subtle faults, and that complex specifications can exacerbate this problem.


%----------------------------------------------------------------------------
\subsubsection{\AutoProof}
%----------------------------------------------------------------------------

\AutoProof is an auto-active verifier of full functional correctness and has been described in Chapter~\ref{sec:ap}.
\AutoProof supports some advanced language constructs such as function objects (agents in Eiffel terminology). Nonetheless, some features of Eiffel---most notably exceptions and floating point arithmetic---are still unsupported and routines using them are not adequately translated to Boogie.
The performance of \AutoProof depends on the quality of contracts available; accurate contracts improve the modularity of the analysis which can then also verify partial implementations.


%----------------------------------------------------------------------------
\subsubsection{\Inspector}
%----------------------------------------------------------------------------

The \Inspector~\cite{ZURFLUH14} is a light-weight code analysis tool to find code smells and simple coding errors along the lines of PMD for Java~\cite{PMD} or FxCop for C\#~\cite{FXCOP}. The \Inspector has an extendible rule system and a large set of implemented rules focusing on code smells like overly complex implementations or simplifiable expressions and also a few coding errors like wrongly used loop iterators. The analysis is in general quick and supports the full Eiffel language.


