\chapter*{Acknowledgments}
%\chapterimage{images/acknowledgments.jpg}


%1. Bertrand Meyer

I thank Bertrand Meyer for giving me the opportunity to do a PhD at ETH Zurich.
His support and the programming language he created were the foundation of this work.

%2. Martin Nordio and Carlo A. Furia

Also, this dissertation would not have been possible without the help and supervision of Martin Nordio and Carlo A. Furia.
Their expertise and feedback was invaluable for my progress.

%3. Nadia Polikarpova and Christian Estler

A direct impact to this work also comes from Nadia Polikarpova, who designed an awesome framing methodology and was a co-developer of \AutoProof during its recent stages, and Christian Estler, who put \AutoProof on the web.

%4. Co-supervisors

Additionally I thank my co-examiners Marsha Chechik, Carlo A. Furia, Francesco Logozzo, and Jim Woodcock for evaluating my dissertation.

%5. Coworkers

My time in this research group was very enjoyable thanks to all the wonderful co-workers:
 Andrey Rusakov,
 Benjamin Morandi,
 Carlo A. Furia,
 Chandrakana Nandi,
 Christian Estler,
 Claudia G\"unthart,
 Cristiano Cal\-ca\-gno,
 \DH urica Nikoli\'{c},
 Georgiana Caltais,
 Jason (Yi) Wei,
 Jiwon Shin,
 Marco Piccioni,
 Marco Trudel,
 Martin Nordio,
 Max (Yu) Pei,
 Michela Pedroni,
 Mischael Schill,
 Roman Schmocker,
 Scott West,
 Sebastian Nanz,
 Stephan van Staden.

%6. People at Eiffel Software

I also thank Manu Stapf from Eiffel Software, who speedily replied whenever I had a question about EiffelStudio.

%7. Family and Friends

During my studies at ETH and before I found wonderful friends with whom I had and continue to have a great time: 
 Antoni Amerena,
 Antonio Barresi,
 Till Bay,
 Nicola Chiapolini,
 Urs D\"onni,
 Bruno Hauser,
 Sven Mader,
 Lukas Naef,
 Tahmineh Sanamrad,
 Martin Seiler,
 Valentin W\"ustholz.

Lastly I thank my family, who supported me on my journey all my life, and in particular my wife Maria Gubar (aka \emph{LimKis}), who designed the amazing cover page and illustrations.


